---
title: "A Temporal Look At Californias Regional Shortage Levels"
author: "Takeen Shamloo"
format: html
editor: source
---

In this Quarto document we will be taking a look at `actual` & `forecast` monthly shortage values grouped by year. Our `actual` shortage values range between *2022-2024* while our `forecast` values are from a range between *2024-2026*. 

Assuming the necessary pre-processing has already been completed, the data is being accessed through a duckdb connection to our sql database. We will be running queries in order to create visuals for some of the organizations available. 

Please review the following files for previous steps:
  initial_data_cleaning.R - data initialization and cleaning.
  shortage_db.sql - loading data into tables and performing analysis via sub-tables.
  shortage_db.duckdb - db file.
  **OPTIONAL** data/results/combined_yearly_avg.csv - result table used below but output as csv.
  
**Key to understanding shortage levels:**

0 = No Shortage
1 : < 10% Shortage 
2 : 10-19% Shortage
3 : 20-29% Shortage
4 : 30-39% Shortage
5 : 40-49% Shortage 
6 : Greater than 50% Shortage

**For this assignment our goal is to answer the following question:**

## How have water shortage levels in California changed over time, and are they reasonably forecasted through time?

```{r}
# Load packages.
library(DBI)
library(duckdb)
library(ggplot2)
library(tools)

# Initialzie database connection to previously created .duckdb file.
connection_to_db <- dbConnect(duckdb::duckdb(), "shortage_db.duckdb")
```

### Example 1: Organization 193 (Beverly Hills)

```{r}
# Selecting the following variables from our combined_yearly_avg only for org_id = 193.
# Ordering by year / data_type.
org193 <- dbGetQuery(connection_to_db, "
    SELECT supplier_name, year, data_type, avg_shortage_level
    FROM combined_yearly_avg 
    WHERE org_id = 193
    ORDER BY year, data_type
")

# View query table.
print(org193)

# Clean up org supplier name for plot title.
org_name <- toTitleCase(unique(org193$supplier_name))

# Plot data for organization grouped by data_type (i.e. actual/forecasted).
ggplot(org193, aes(x = year, y = avg_shortage_level, color = data_type, group = data_type)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    labs(title = paste('Water Shortage Levels for', org_name),
         x = 'Year', 
         y = 'Average Shortage Level', 
         color = 'Data Type') +
    theme_minimal() +
    theme(text = element_text(size = 12))
```
From our graph we can see that Beverly Hills despite carrying a **2-2.5** level
shortage accross *2022-2024* but were forecasted to have a level of **0** between 
*2024-2025*. Additionally no data was forecasted for *2026*.  

### Example 2: Organization 1873 (Pasadena)

```{r}
# Same query as the previous two except for org_id = 1873.
org1873 <- dbGetQuery(connection_to_db, "
    SELECT supplier_name, year, data_type, avg_shortage_level
    FROM combined_yearly_avg 
    WHERE org_id = 1873
    ORDER BY year, data_type
")

# View query table.
print(org1873)

# Clean up org supplier name for plot title.
org_name <- toTitleCase(unique(org1873$supplier_name))

# Plot data for organization grouped by data_type (i.e. actual/forecasted).
ggplot(org1873, aes(x = year, y = avg_shortage_level, color = data_type, group = data_type)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    labs(title = paste('Water Shortage Levels for', org_name),
         x = 'Year', 
         y = 'Average Shortage Level', 
         color = 'Data Type') +
    theme_minimal() +
    theme(text = element_text(size = 12))
```

From our graph we can see that the City of Pasadena despite carrying a 
**2-2.25** level shortage accross *2022-2024* but were forecasted to have a level 
of **0** between *2025-2026*. We should note that there was no values forecasted
for *2024* therefore we see a gap in our data.

### Example 3: Organization 23 (Alameda County Water District)

```{r}
# Same query as the previous two except for org_id = 23.
org23 <- dbGetQuery(connection_to_db, "
    SELECT supplier_name, year, data_type, avg_shortage_level
    FROM combined_yearly_avg 
    WHERE org_id = 23
    ORDER BY year, data_type
")

# View query table.
print(org23)

# Clean up org supplier name for plot title.
org_name <- toTitleCase(unique(org23$supplier_name))

# Plot data for organization grouped by data_type (i.e. actual/forecasted).
ggplot(org23, aes(x = year, y = avg_shortage_level, color = data_type, group = data_type)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    labs(title = paste('Water Shortage Levels for', org_name),
         x = 'Year', 
         y = 'Average Shortage Level', 
         color = 'Data Type') +
    theme_minimal() +
    theme(text = element_text(size = 12))
```

From our graph we can see that the Alameda County Water District despite 
carrying showing a trend where the shortage level falls from **2** in *2022*
all the way until **0** in *2024*. They were forecasted to have a level of 
**~1.25-1** between *2024-2025*. Additionally no data was forecasted for *2026*.  

### Example 4: Organization 1238 (Inglewood)

```{r}
# Same query as the previous two except for org_id = 1238.
org1238 <- dbGetQuery(connection_to_db, "
    SELECT supplier_name, year, data_type, avg_shortage_level
    FROM combined_yearly_avg 
    WHERE org_id = 1238
    ORDER BY year, data_type
")

# View query table.
print(org1238)

# Clean up org supplier name for plot title.
org_name <- toTitleCase(unique(org1238$supplier_name))

# Plot data for organization grouped by data_type (i.e. actual/forecasted).
ggplot(org1238, aes(x = year, y = avg_shortage_level, color = data_type, group = data_type)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    labs(title = paste('Water Shortage Levels for', org_name),
         x = 'Year', 
         y = 'Average Shortage Level', 
         color = 'Data Type') +
    theme_minimal() +
    theme(text = element_text(size = 12))
```

From our graph we can see that the City of Inglewood despite showing the highest
of all the other examples beginning from a shortage level of **3** level in
*2022* increasing to a **4** in *2024* but where forecasted to have a level of
**0** between *2024-2025*. Additionally no data was forecasted for *2026*.  

#### Important! Make sure to shut down db.
```{r}
dbDisconnect(connection_to_db, shutdown = TRUE)
```
